Adaptive evolution with aneuploidy and mutation

### 2022-06-01 ms/ms.tex
revised new results section

\paragraph*{Genotype frequency dynamics.}
Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\Cref{fig:ppc-plot}). % TODO new figure by Martin and discussion of the probability of 2n* arriving via 2n+1 - transient solution etc.

% Fig dynamics
%% generated with dynamics-plot.ipynb
\begin{figure}[h!]
  \centering
\includegraphics[width=0.8\textwidth]{../figures/dynamics.pdf}
  \caption{
  \textbf{Posterior genotype frequency dynamics for the model.} 
  The posterior prediction for the frequencies of the four genotypes over time. Each of the 20 curves is the average of 10,000 simulations of the model using parameters drawn from the posterior distribution.
  }
  \label{fig:ppc-plot}
\end{figure}

>>>>

\paragraph*{Competition between mutation and aneuploidy.}

We simulated 50 % TODO is it 50?
 replicate genotype frequency dynamics using the MAP estimate parameters. 
\Cref{fig:ppc-plot}A shows the simulated frequencies of the four genotypes (\euwt, \anwt, \anmt\ and \eumt), as well as the frequencies of \eumt\ cells that arose from either \anwt\ cells via a sequences of mutation and chromosome loss events, or directly from \euwt\ cells via a mutation event.
We find that \anmt\ never reaches substantial frequency as it is quickly replaced by \eumt.

\citet{Yona2012} have suggested that aneuploidy is a transient adaptive solution, arising quickly due to increased fitness and high formation rate, but 
ultimately lost as beneficial mutations appear and the extra chromosome is lost due to the cost of aneuploidy. 
We tested this scenario by estimating $F_A$, the expected frequency of \eumt\ that arose from \anwt\ (as the average frequency of such cells at the end of 50 replicate simulations).
Surprisingly, we observe that the majority ($F_A=XX$) % TODO add number
of \eumt\ cells are a product of a direct mutation in \euwt\ cells, rather than descending from \anwt\ cells. 
This is despite the fact that the \anwt\ genotype reaches high frequencies in the population (\Cref{fig:ppc-plot}A). % TODO add number
This result is not unique to the MAP parameter estimate. 
We simulated genotype frequency dynamics using parameter samples from the posterior distribution (\Cref{fig:posterior}), and computed the posterior distribution of $F_A$ (\Cref{fig:ppc-plot}B). 
The mean $F_A$ was just XX [XX-XX 95\% CI], % TODO add number
such that if we pick up a cell from the evolved population it is more likely to have descended directly from a euploid cell than from an aneuploid cell.
As expected, the probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate, $\mu$ (\Cref{fig:ppc-plot}C and D).

Only in 681 of 100,000 posterior samples (0.6\%) $F_A$ was  larger than 0.5, that is, \eumt\ was more likely to occur via aneuploidy lineage.  
These parameter samples can be roughly divided to two clusters (\Cref{fig:ppc-plot}C and E).
In the first, aneuploidy formation is not much faster than mutation ($\delta/\mu \approx 500$), but this is compensated by a small fitness difference between \eumt\ and \anwt\ ($w_{\eumt}/w_{\anwt}-1 \approx 0.0055$), which results in a low fixation probability for \eumt\ once \anwt\ is at a high frequency, and limits the strength of clonal interference.
In the second cluster, aneuploidy formation is much faster than mutation ($\delta/\mu \approx 10^5$), but the fitness difference between \eumt\ and \anwt\ is also larger ($w_{\eumt}/w_{\anwt}-1 \approx 0.007$). 
Thus, once \eumt\ appears it more easily avoids extinction and competes with \anwt. This is compensated by the high aneuploidy rate (note that we only use samples from the posterior, that is, we only compute $F_A$ for parameter choices that reasonably agree with the experimental results.)

% Fig dynamics
\begin{figure}[h!]
  \begin{subfigure}{0.5\textwidth}
      \centering
      %% generated with dynamics-plot.ipynb
      \includegraphics[width=\textwidth]{../figures/dynamics.pdf}      
  \end{subfigure}
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/histo_100000.pdf}
  \end{subfigure}
  \\
  \begin{subfigure}{0.33\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/scatter100000_log_star.png}
  \end{subfigure}
  \begin{subfigure}{0.33\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/scatter_dirmut_log_star_byr.png}
  \end{subfigure}
  \begin{subfigure}{0.33\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/scatter100000_rater_fitr.png}
  \end{subfigure}
  \caption{
    \textbf{Posterior genotype frequency dynamics.}     
    \textbf{(A)} The posterior prediction for the frequencies of the four genotypes over time, including the source of \eumt. Each of the curves is the average of 50 simulations  using parameters drawn from the posterior distribution. % TODO check if true
    \textbf{(B)} Marginal posterior distribution of $F_A$, the expected frequency of \eumt\ cells that arose from \anwt cells, computed as the average frequency at the end of 50 simulations for each of 100,000 samples from the parameter posterior distribution. Solid and dashed lines show the mean and 95\% CI.
	\textbf{(C-E)} $F_A$ values (color coded) from panel B, with their corresponding parameter values on the x- and y-axes. White start shows the MAP estimate. Panel D only shows $F_A<0.5$.
     }
  \label{fig:ppc-plot}
\end{figure}
